Transparency to hybrid open access invoicing through publisher-provided metadata: A longitudinal article-level study of Elsevier

### 2021-01-29 analysis/manuscript.tex
version pushed to overleaf for final polishing

\title{Your APA6-Style Manuscript}
\shorttitle{}
\author{You}
\affiliation{Somewhere}

\abstract{Your abstract here.}

\begin{document}

@@

The problem with this lack of transparency becomes more apparent in the
case of hybrid OA. In this OA business model, individual articles can be
made openly available for a fee, also known as article-processing charge
(APC), while the journal as a whole remains subscription access. Hence,
hybrid OA was originally introduced as a low-risk transitional model
that allows journals to gradually convert to full OA and reduce
subscription costs as the OA uptake increases (Prosser, 2003). Many
large subscription publishers have since incorporated this OA model, and
while the uptake among authors increased only slowly during the first
decade (Björk, 2012), it has gradually gained popularity in recent
years. This development was driven by the implementation of OA policies
by research institutions and funders, the allocation of OA funds, and
formal agreements between publishers and institutions that allow
affiliated authors to publish OA free-of-charge. However, there have
been concerns that the hybrid model allows for
double-dipping---receiving two payments for one article, the APC and
subscription fees (Prosser, 2015; Shieber, 2009). While publishers
assure they combine subscription fees and APCs to adjust pricing
(Mittermaier, 2015), without transparency around the uptake of hybrid OA
and revenue streams, such claims are impossible to evaluate.

With the recent introduction of transformative agreements, an evolving
term for contracts to shift library spendings for subscriptions to OA
(Borrego et al., 2020; Hinchliffe, 2019), the demand for
publisher-provided data has even increased. If publishers would enable
transparency about the OA uptake in hybrid journals and APC invoicing,
academic libraries could improve their assessments and adjustments of
contracts with publishers (Antelman, 2017). Likewise, research
organizations and funders could improve compliance monitoring with OA
mandates, while ensuring that the different actors were not charged
multiple times (Larivière \& Sugimoto, 2018). However, as formal
bibliometrics and financial studies have noted, the lack of publicly
available and standardized data about hybrid OA publishing is
considerable (Laakso \& Björk, 2016; Lawson, 2015; Pinfield et al.,
2016)

This current study investigates the extent to which publisher-provided
data can be utilized to assess hybrid OA. The study takes Elsevier, the
largest scholarly journal publisher in terms of article volume
(Larivière et al., 2015), as the starting point for an empirical
analysis of hybrid OA uptake and invoicing. Our approach is based on
openly available data sources and brings together scholarly data from
Crossref and machine-readable APC invoicing data from Elsevier OA
full-texts, which enables us to analyze the development of Elsevier's
hybrid journal portfolio from 2015-19, and to compare it with that of
Elsevier's full and delayed OA program. Further, we investigate whether
hybrid publishing fees were waived or charged to the authors or academic
institutions or research funders with Elsevier OA agreements. In
particular, we identify academic consortia and research funders that
were directly invoiced by Elsevier, and compare their hybrid OA rates by
license and subject. As such, our findings have implications for
research and OA policy and highlight the potential of publisher-provided
data for large-scale, instantaneous analysis of hybrid OA uptake and
invoicing.

@@

Other studies assessed the share of (hybrid) OA for the entire corpus of
scholarly research articles. For instance, using research articles with
an assigned DOI and indexed in the Web of Science Core Collection
(n=2,610,305) as a benchmark, Martín-Martín et al. (2018) assessed how
many research articles from 2009 and 2014 are freely available through
Google Scholar. After identifying OA articles through licensing
information from Crossref, the authors estimated the share of hybrid OA
at around 0.5\% in 2009 and 1.5\% in 2014. In a similar vein, Piwowar et
al. (2018) drew on Crossref to identify content licenses. The authors
followed a data-driven approach based on Unpaywall data, which is openly
available and contained 66 million Crossref DOIs at the time, to
determine how much of the scholarly literature is OA. Based on a random
sample of 100,000 Crossref DOIs, the authors found 3.6\% of hybrid OA,
while pointing towards a growing trend in recent years that recorded a
hybrid OA share of 9.4\% of all articles published in 2015. While
Piwowar et al.'s (2018) and Martín-Martín et al.'s (2018) methods are
well-suited for large-scale and repeated analyses of the prevalence of
hybrid OA, their main limitation is that they lack information about
publishing costs.

Moreover, Unpaywall's hybrid OA detection method has been refined
several times since its launch in 2017, reflecting growing albeit
incomprehensive license information in Crossref metadata. To overcome
these shortcomings, the developers extended previous web-scraping
approaches to better distinguish between immediate and delayed OA based
on licenses at the date of observation Piwowar et al. (2019). Relatedly,
in October 2020, Elsevier's license for delayed OA content was added to
Unpaywall's list of license patterns (Unpaywall, n.d.-c). Finally, at
the journal-level, a more fine-grained identification of full OA
journals using OA article information was developed to better
distinguish them from subscription journals including hybrid journals
(Unpaywall, n.d.-b). Because of these ongoing changes due to
non-standardized data, uptake studies reporting hybrid OA uptake based
on Unpaywall data cannot easily be compared with each other.

@@

As this brief review has shown, the most comprehensive studies approach
hybrid OA funding from an institutional perspective that focuses on
research institutions or funders and draws on internal records or
crowd-sourced data like from the Open APC initiative. The advantages of
Open APC data are that they reflect the actual APCs paid rather than
list prices; they are not limited to certain publishers or journals;
they are provided in a standardized format regardless of the publisher.
On the other hand, the drawbacks include the limited number of
participating institutions that are almost exclusively based in Europe
and the dependency on participating organizations for data integrity.
Moreover, since Open APC data does not include waivers or free APCs, it
cannot present a complete picture of all publishing activities and
expenses. This, however, would pose less of an issue when analyzing just
hybrid OA.

\hypertarget{data-driven-workflows-for-hybrid-oa}{%
\subsection{Data-driven workflows for hybrid
OA}\label{data-driven-workflows-for-hybrid-oa}}

In September 2018, coalition S presented a set of principles known as
Plan S and announced. that from 2021 onwards, signatories would no
longer fund publications in hybrid OA journals. This principle includes
so-called ``mirror'' and sister journals---hybrid journals with fully OA
counterparts or split into their fully OA and subscription parts---that
have launched since Plan S was announced but makes an exception for
hybrid journals with transformative agreements (Frantsvåg \& Strømme,
2019). While the scale and means of these transformative agreements may
differ, many address the lack of transparency surrounding hybrid OA by
making publisher-provided metadata a defining element of publishing
workflows between institutions and publishers (Geschuhn \& Stone, 2017;
Marques et al., 2019)

Elsevier, one of the world's largest scholarly publishers, provides
article-level metadata to Crossref, including license information, which
can be used to identify OA articles, and links to HTML and XML full-text
versions for text and data mining purposes. Surprisingly, besides the
full-text article, the XML files also contain invoicing information and
the article's OA status. Table \ref{tab:els_xml} summarizes the key
metadata elements, which comprise the type of entity to which the APC
was invoiced (\texttt{openaccessSponsorType}) and the name of that
entity if available (\texttt{openaccessSponsorName}). Moreover, the
article-level metadata indicates if OA was provided through Elsevier's
Open Archive program (\texttt{openArchiveArticle}), which enables free
access to participating subscription journals after an embargo period
(i.e., delayed OA; Elsevier (n.d.-b)). As we have been unable to
identify similar efforts by other large publishers (e.g., Springer
Nature), this increased transparency through openly available
bibliometric and financial data at the article level seems unique to
Elsevier.

\begin{table}[H]
\caption{\label{tab:els_xml}Overview of hybrid open access (OA) uptake across Elsevier hybrid journals with at least one OA article 2015-2019.}
\centering
\begin{tabular}[t]{lp{10cm}}
\toprule
XML Node & Description\\
\midrule
\texttt{openaccessArticle} & Indicates if the article is open
access. \\
\texttt{openaccessSponsorType} & Invoice recipient type.\\
\texttt{openaccessSponsorName} & Entity that received Elsevier's APC
invoice.\\
\texttt{openArchiveArticle} & Indicates if open access was provided through
Elsevier's Open Archive programme.\tabularnewline \\
\bottomrule
\end{tabular}
\end{table}

Table 1: Overview of metadata provided in XML-based Elsevier full-texts
representing information about open access status and invoicing. A
sample XML is available here:
\url{https://api.elsevier.com/content/article/PII:S0169409X18301479?httpAccept=text/xml}

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

>>>>

\title{Transparency to hybrid open access invoicing through publisher-provided metadata: A longitudinal article-level study of Elsevier}
\shorttitle{Hybrid OA Elsevier}
% \author{Najko Jahn}
% \affiliation{State and University Library Göttingen, University of Göttingen, Germany}
% \author{Lisa Matthias}
% \affiliation{tba}
% \author{Mikael Laakso}
% \affiliation{tba}
%

\abstract{The financial flows in scholarly journal publishing have become increasingly complex with the growth of open access (OA), and the lack of data and transparency into these flows has been persistent. Opaqueness has concerned especially hybrid OA, where subscription-based journals also publish OA content if an optional fee is paid. By leveraging Elsevier’s recently introduced article metadata concerning invoiced recipients, this study provides the first longitudinal publisher-level study of uptake and invoicing of hybrid OA articles. The results show that the share of OA articles in Elseviers hybrid journals has grown slowly, from 2.6\% of all articles on average in 2015 to 3.7\% in 2019. About a third of hybrid OA articles are invoiced directly through an agreement with research funders or other type of organisational sponsor (e.g. national library consortia), with the majority invoiced to authors. 83\% of agreement-invoiced hybrid OA articles are to European organisations, suggesting a strong international skewness. The study discovered a large discrepancy in choice of publication license between author-invoiced articles (dominantly CC BY-NC-ND) and agreement-invoiced articles (dominantly CC BY). The results demonstrate the value of publisher-provided metadata covering invoiced recipients of articles, which would be even more useful through wider adoption among publishers.}

\begin{document}
\maketitle

@@

The problem with this lack of transparency becomes more apparent in the
case of hybrid OA. In this OA business model, individual articles can be
made openly available for a fee, also known as article-processing charge
(APC), while the journal as a whole remains subscription access. Hence,
hybrid OA was originally introduced as a low-risk transitional model
that allows journals to gradually convert to full OA and reduce
subscription costs as the OA uptake increases (Prosser, 2003). Many
large subscription publishers have since incorporated this OA model, and
while the uptake among authors increased only slowly during the first
decade (Björk, 2012), it has gradually gained popularity in recent
years. This development was driven by the implementation of OA policies
by research institutions and funders, the allocation of OA funds, and
formal agreements between publishers and institutions that allow
affiliated authors to publish OA free-of-charge (Laakso \& Björk, 2016).
However, there have been concerns that the hybrid model allows for
double-dipping---receiving two payments for one article, the APC and
subscription fees (Prosser, 2015; Shieber, 2009). While publishers
assure they combine subscription fees and APCs to adjust pricing
(Mittermaier, 2015), without transparency around the uptake of hybrid OA
and revenue streams, such claims are impossible to evaluate.

With the recent introduction of transformative agreements, an evolving
term for contracts that shift library spendings for subscriptions to OA
(Borrego et al., 2020; Hinchliffe, 2019), the demand for
publisher-provided data has even increased. If publishers would enable
transparency about the OA uptake in hybrid journals and APC invoicing,
academic libraries could improve their assessments and adjustments of
contracts with publishers (Antelman, 2017; Schimmer et al., 2015).
Likewise, research organizations and funders could improve compliance
monitoring with OA mandates, while ensuring that the different actors
were not charged multiple times (Larivière \& Sugimoto, 2018). However,
as formal bibliometrics and financial studies have noted, the lack of
publicly available and standardized data about hybrid OA publishing is
considerable (Laakso \& Björk, 2016; Lawson, 2015; Pinfield et al.,
2016)

This current study investigates the extent to which publisher-provided
data can be utilized to research hybrid OA. The study takes Elsevier,
the largest scholarly journal publisher in terms of article volume
(Larivière et al., 2015), as the starting point for an empirical
analysis of hybrid OA uptake and invoicing. Our approach is based on
openly available data sources and brings together scholarly metadata
from Crossref, a DOI registration agency for scholarly works, and
machine-readable APC invoicing data from Elsevier OA full-texts. The
so-compiled data enable us to analyze the development of Elsevier's
hybrid journal portfolio from 2015-19, and to compare it with that of
Elsevier's full and delayed OA program. Further, we investigate whether
hybrid publishing fees were waived or charged. In particular, we
identify academic consortia and research funders that were directly
invoiced by Elsevier, and compare their hybrid OA rates by license and
subject. As such, our findings have implications for research and OA
policy and highlight the potential of publisher-provided data for
large-scale, instantaneous analysis of hybrid OA uptake and invoicing.

@@

Other studies assessed the share of (hybrid) OA for the entire corpus of
scholarly research articles. For instance, using research articles with
an assigned DOI and indexed in the Web of Science Core Collection
(n=2,610,305) as a benchmark, Martín-Martín et al. (2018) assessed how
many research articles from 2009 and 2014 are freely available through
Google Scholar. After identifying OA articles through licensing
information from Crossref, the authors estimated the share of hybrid OA
at around 0.5\% in 2009 and 1.5\% in 2014. In a similar vein, Piwowar et
al. (2018) drew on Crossref to identify content licenses. The authors
followed a data-driven approach based on Unpaywall data, which is openly
available and contained 66 million Crossref DOIs at the time, to
determine how much of the scholarly literature is OA. Based on a random
sample of 100,000 Crossref DOIs, the authors found 3.6\% of hybrid OA,
while pointing towards a growing trend in recent years that recorded a
hybrid OA share of 9.4\% of all articles published in 2015.

Since its launch, Unpaywall has become widely used as an OA evidence
source in bibliometric research and rankings(Huang et al., 2020;
Robinson-Garcia et al., 2020). However, results cannot easily be
compared with each other over time because of Unpaywall's ongoing
changes due to non-standardized data to discover hybrid OA. These
changes mainly reflect the growing albeit incomprehensive license
information in Crossref metadata. To overcome these shortcomings, the
developers extended previous web-scraping approaches to better
distinguish between immediate and delayed OA based on licenses at the
date of observation Piwowar et al. (2019). Relatedly, in October 2020,
Elsevier's license for delayed OA content was added to Unpaywall's list
of license patterns (Unpaywall, n.d.-c). At the journal-level, a more
fine-grained identification of full OA journals using OA article
information was developed to better distinguish them from subscription
journals including hybrid journals (Unpaywall, n.d.-b). With its servie
unsub, Unpaywall recently launched a paid service, where libraries can
compare hybrid OA uptake rates with internal financial and usage data to
assess the value of journal subscriptions (Chawla, 2020).

@@

As this brief review has shown, the most comprehensive studies approach
hybrid OA funding from an institutional perspective that focuses on
research institutions or funders and draws on internal records or
crowd-sourced data like from the Open APC initiative. The advantages of
Open APC data are that they reflect the actual APCs paid rather than
list prices; they are not limited to certain publishers or journals;
they are provided in a standardized format regardless of the publisher.
On the other hand, the drawbacks include the limited number of
participating institutions that are almost exclusively based in Europe
and the dependency on participating organizations for data integrity
(Pieper \& Broschinski, 2018). Moreover, since Open APC data does not
include waivers or free APCs, it cannot present a complete picture of
all publishing activities and expenses. This, however, would pose less
of an issue when analyzing just hybrid OA.

\hypertarget{the-role-of-publisher-provided-data}{%
\subsection{The role of publisher-provided
data}\label{the-role-of-publisher-provided-data}}

Uptake and financial studies have OA policy implications for hybrid OA.
Most prominently, the cOAlition S, an international network of research
funders, released its widely discussed Plan S. In September 2018, the
cOAlition S announced that from 2021 onwards, signatories would no
longer fund articles in hybrid OA journals. This principle includes
so-called ``mirror'' and sister journals--hybrid journals with fully OA
counterparts or split into their fully OA and subscription parts---that
have launched since Plan S was announced but makes an exception for
hybrid journals with transformative agreements (Frantsvåg \& Strømme,
2019). The Plan S refers to the Efficiency and Standards for Article
Charges (ESAC) guidelines (Geschuhn \& Stone, 2017) as critical
component of transformative agreements, emphasizing the need for
publisher-provided metadata to address the lack of transparency
surrounding hybrid OA. However, studies confirm that publisher with such
agreements lack essential data about hybrid OA uptake and invoicing, or
are not able or willing to provide such information publicly and
instantaneously (Marques et al., 2019; Monaghan et al., 2020), which, in
turn, not just limit the monitoring of hybrid OA, but also research
studies on its size and financial implications.

Here, we focus on Elsevier. The publisher provides article-level
metadata to Crossref, including license information, which can be used
to identify OA articles, and links to HTML and XML full-text versions
for text and data mining purposes. Surprisingly, besides the full-text
article, the XML files also contain invoicing information and the
article's OA status. Table \ref{tab:els_xml} summarizes the key metadata
elements, which comprise the type of entity to which the APC was
invoiced (\texttt{openaccessSponsorType}) and the name of that entity if
available (\texttt{openaccessSponsorName}). Moreover, the article-level
metadata indicates if OA was provided through Elsevier's Open Archive
program (\texttt{openArchiveArticle}), which enables free access to
participating subscription journals after an embargo period (i.e.,
delayed OA; Elsevier (n.d.-b)). As we have been unable to identify
similar efforts by other large publishers (e.g., Springer Nature), this
increased transparency through openly available bibliometric and
financial data at the article level seems unique to Elsevier.

\begin{table}[H]
\caption{\label{tab:els_xml} Overview of hybrid open access (OA) uptake across Elsevier hybrid journals with at least one OA article 2015-2019}
\centering
\begin{tabular}[t]{lp{10cm}}
\toprule
XML Node & Description\\
\midrule
\texttt{openaccessArticle} & Indicates if the article is open
access. \\
\texttt{openaccessSponsorType} & Invoice recipient type.\\
\texttt{openaccessSponsorName} & Entity that received Elsevier's APC
invoice.\\
\texttt{openArchiveArticle} & Indicates if open access was provided through
Elsevier's Open Archive programme.\tabularnewline \\
\bottomrule
\end{tabular}
\end{table}

@@

\hypertarget{discussion}{%
\section{Discussion}\label{discussion}}

Over the last five years, Elsevier recorded growth in the uptake of
hybrid OA for each of our three indicators: The number of hybrid OA
articles published per year doubled, the number of hybrid journals with
at least one OA article grew by 21\%, and the share of hybrid OA
articles relative to closed-accessed articles in these journals
increased from 2.6\% to 3.7\%. As Laakso \& Björk (2016), we found only
a weak relationship between citation metrics like the SNIP and hybrid OA
uptake and observed differences in the uptake of hybrid OA among journal
subject areas. In particular, we found the highest count of hybrid OA
articles in physical sciences journals (see Table
\ref{tab:subject_area_table}). This was followed by the life sciences
and health sciences, whereas the social sciences had the lowest count of
hybrid articles. This order mostly reflects the disciplines' overall
publication output. According to the Open Science Monitor by the
European Commission, the physical sciences and mathematics publish the
most articles, followed by the health sciences, life sciences, and
social sciences and humanities.

Such disciplinary differences in hybrid OA prevalence become more
meaningful when considering the relative share of OA articles to
closed-access articles. In line with previous research, we found that
Elsevier journals from the life sciences and social sciences (Jubb et
al., 2017; Kramer \& Bosman, 2018; Laakso \& Björk, 2016) recorded
greater than typical hybrid OA uptake, whereas physical sciences
journals generally had a lower than typical uptake (Kramer \& Bosman,
2018; Laakso \& Björk, 2016; Martín-Martín et al., 2018). In a
systematic review of OA publishing patterns, Severin et al. (2020)
linked such disciplinary differences to socio-cultural and technological
factors that shape publishing cultures and practices. Among other
aspects, the authors pointed to the availability of prestigious
journals, the possibility to provide OA through alternative routes, and
ease of access to funding sources to cover publishing fees. As such,
looking at hybrid OA uptake in isolation can only provide limited
insight, so that placing it within the broader context of disciplinary
publishing and OA practices is necessary. For instance, since many
branches of the physical sciences have established self-archiving
practices and the technical infrastructure to support these (e.g., the
arXiv; Severin et al. (2020), Björk et al. (2010), Laakso \& Björk
(2016)), researchers can provide OA through repositories (i.e., green
OA), and indeed, according to the Open Science Monitor the physical
sciences provide OA primarily through the green route (31.2\% total OA
share, 24\% OA if only considering green OA). Hence, the low share of
hybrid OA in our sample might be because physical scientists perceive
much less of a need for hybrid OA. Further, Severin et al. (2020) note
project-based funding structures enable OA publishing in fully OA and
hybrid journals as they allow for easy integration of publishing costs.
Since these funding structures are typical for the life sciences
(ibid.), we would expect higher rates of OA. Data from the Open Science
Monitor seems to support this notion as the life sciences record
particularly high shares of full OA articles and comparatively high
shares of hybrid OA in some fields, such as biology and agricultural
science. While we expected the observed hybrid OA shares for the life
sciences and physical sciences, we were surprised by the high share of
hybrid OA belonging to the AJSC subject area social sciences, which also
includes the humanities, since articles ``social scientists have
reported to face significant difficulties in access to grant funding for
{[}publication charges{]}, as most research in the social sciences is
not done by means of project-specific funding'' Severin et al. (2020, p.
15). However, previous research has found comparatively low awareness
and uptake of green OA among researchers in the field (Björk et al.,
2010; Creaser et al., 2010), which might encourage hybrid OA uptake.
Another possible explanation,is that Elsevier has a number of high
prestigious social science and humanities journals (Larivière et al.,
2015), which means that our findings are not representative of the
general population of SSH journals. Likewise, this limitation could be
extended to our findings in general that whilst not representative of
the population of hybrid OA publishing, they provide valuable insight
into a key player of scholarly publishing.

Finally, OA policies in favor of hybrid OA in combination with invoicing
agreements with Elsevier might be a driver. For instance, we found that
nearly half of articles in arts and humanities were facilitated through
agreements, which was above-average. This is in line with previous
research, suggesting that OA policies and funder mandates have a
stronger effect on OA uptake than disciplinary publication cultures
(Huang et al., 2020; Larivière \& Sugimoto, 2018).

Our overall investigation of invoice channels and recipients of hybrid
OA APCs found that the majority of Elsevier hybrid articles were
invoiced to the author (n=41,725; 58.2\%). However, it is important to
note that although Elsevier's metadata classifies the sponsor type as
``author'', this does not necessarily mean that authors paid for APCs
out of their own pocket but rather just that the APC was not invoiced
through an agreement. Hence, it is possible that APCs invoiced to the
author were covered by institutional funds or as part of research
grants. For instance, a recent Springer Nature survey found that hybrid
OA was predominantly supported through institutional and funder sources
(71\%), followed by OA agreements (34\%), while only 6\% were paid from
personal funds or savings (Monaghan et al., 2020). We also observed
notable differences in licensing among invoice types. The majority of
hybrid OA articles, regardless of the invoice type, were licensed under
the more restrictive CC BY-NC-ND license. There is a lack of dedicated
studies on licence selection, but indications so far suggest that when
given a choice authors tend to select more restrictive variants for
their published works (Fraser et al., 2020; Noorden, 2013; Rowley et
al., 2017). On the other hand, when hybrid articles were invoiced
through agreements, most articles were licensed under the more liberal
CC BY license. As several funding bodies mandate CC BY licenses,
including the UK research funders that account for 62\% of our
``agreement'' subsample, this result is perhaps not surprising but
suggests the effectiveness of such agreements.

Based on our findings, around a third of the articles were invoiced
directly through OA agreements (e.g., research funders, national library
consortia). Strikingly, sponsors based in the UK accounted for more than
half of these , and a quarter was charged to sponsors from other
European countries. This finding reiterates reports from previous
studies that the UK's OA profile differs markedly from other countries
(cf.~ref, ref) and points to the effects of science policy. To increase
OA to research publications, the UK implemented centralized APC funding
and embraced hybrid OA as a transition model. Within five years of the
publication of the Finch Report in 2012, the UK recorded an 18\%
increase in immediate OA, coupled with a rise in hybrid OA from 2.7\% in
2012 to 15.4\% of all articles in 2016 (Jubb et al., 2017), which can be
attributed to the availability of APC funding from Research Councils UK
(RCUK). Indeed, RCUK block grants have been the largest single source of
OA funds in the UK---the Wellcome Trust, and more recently
transformative agreements (Jubb et al., 2017; Tickel, 2018). However,
since then publishing expenditures of UK universities have been rising
rapidly (ref), so that several universities stopped supporting hybrid OA
in late 2018-2019 (Castro, 2019; University of Birmingham, n.d.;
University of Oxford, 2019; Walker, 2019). This development might
explain the slight decrease in hybrid articles invoiced to the UK our
data showed around that time (see Figure
\ref{fig:invoice_sponsor_country}).

An essential part in the absence of transparency in scholarly publishing
and especially in hybrid OA has been the lack of transparency in
publishers' article management. This has prohibited consumers to get a
comprehensive and comparative overview of their publishing activities
and the associated costs and, thus, limited their negotiating power.
However, certain initiatives in Europe like the ESAC Initiative have
begun to press for improved and more transparent workflows for
transformative agreements (Geschuhn \& Stone, 2017). In particular, the
ESAC Workflow Recommendations seek to increase efficiency and
transparency in the publication workflow through improved invoicing and
reporting processes and metadata delivery, including article-based
machine readable information about the OA sponsor. Through this study we
demonstrated the utility and benefits of such publisher-provided
metadata about hybrid OA funding and highlighted the need for extending
these to include comprehensive information about the licensing and the
particular OA type.

Usually identifying delayed OA articles and distinguishing them from
other types of OA is challenging since there is no publicly available
list of such journals. For instance, Taylor \& Francis' (2021) website
indicates that some of their journals participate in a delayed or
promotional OA program but no further information is provided as to
which journals these are. However, Elsevier provides a list of journals
enrolled in the ``Open Archive'' programme and further tags delayed OA
articles in the machine-readable XML version of the article. This
enables easy identification and more accurate assessments of hybrid OA,
which allows to take into account the substantial amount of delayed OA
content in future hybrid OA studies using Elsevier data.

While this study advances our knowledge about hybrid OA uptake and
invoicing, the limitations leave room for future research. This study
focused on only one publisher, which limits the generalizability of our
findings since Elsevier's attributes such as composition of the journal
portfolio, mix of business models, pricing, and promotion of various
options cannot be assumed to be representative of scholarly journal
publishing in general. Though the invoicing data extracted from the XML
files enables unprecedented accuracy and comprehensiveness in the study
of hybrid OA, it likely does not disclose the actual OA sponsor in all
cases. Most articles were invoiced to authors, but it remains unclear
how often authors themselves pay for APCs or invoices are covered
through institutional OA funds or research grants. Further research into
this topic would increase our understanding of OA funding streams
outside of publishing agreements. Further, our study demonstrates that
to achieve comprehensive mapping and understanding of the financial
flows involved in OA funding, complex, in-depth country-specific
meso-level analyses is needed. Such studies would ideally draw on
various data sources, such as publicly available bibliometrics and
metadata, information about significant national research funders and
their OA policies, information about the national library and possible
consortia landscape including their publishing agreements, which might
require Freedom of Information Requests.

This study provides a snapshot of hybrid OA for the largest journal
publisher prior to the impact of the implementation of Plan S
requirements. While many Plan S signatories have already had strong OA
policies, this harmonized approach is likely to affect publishing
decisions of funded authors,the licencing of their articles, and the
offerings and pricing of publishers. Since the new requirements are
applicable to research funded from 2021 onwards, a comparative study
focused on articles invoiced to Plan S signatories would be a fruitful
endeavor.

In a similar vein, future studies could combine invoice information with
research funding information since these contribute different pieces to
the funding stream puzzle. This could also provide more insight into the
actual APC funding streams behind the author-invoiced articles.

\hypertarget{conclusion}{%
\section{Conclusion}\label{conclusion}}

The primary aim of this empirical study was to investigate Elsevier's
hybrid OA publishing from 2015-2019 to better understand the volume and
invoicing of hybrid OA and present a novel, data-driven approach for
such analyses. Our results indicate that although the number of hybrid
OA articles has increased over time, hybrid OA uptake has remained low.
Notably, hybrid APCs were most often invoiced directly to the authors or
through institutional agreements, where only a few funding bodies were
the main driver of hybrid OA. Finally, our findings highlight that
publisher-provided metadata about the invoicing channels of (hybrid) OA
can facilitate research into and increase the understanding of the
financial flows of OA publishing.

Ever since its inception, hybrid OA has been a challenging subject to
study due to the lack of standardised ways for publishers to flag such
content and APC funding data being limited to studies based on surveys
and articles' acknowledgements sections. This study presented a novel
approach to studying OA APC invoicing that is based on publicly
available publisher-provided metadata, which can be used on on its own
or in combination with other public data sources to gain more detailed
and comprehensive insights into hybrid OA uptake and invoicing than ever
before. If more publishers reported OA invoicing data on the article
level and in a machine-readable format, this would increase transparency
and enable enhanced monitoring of the scholarly journal landscape over
time. As hybrid OA has increasingly become a central element of OA
policies of research funders and libraries, customer organisations could
require that invoicing information is added to the article-level
metadata. As long as publishers do not provide this data in a structured
and comprehensive format, they prevent consumers from benchmarking
prices and therefore hinder price competition as the only alternative to
gain comparative insight is self-reported APC expenditures that research
funders and libraries voluntarily report to the Open APC initiative. The
Open APC initiative is a valuable service but it cannot replace
publisher-provided data since global participation is limited.

From recent science policy developments in Europe it appears that Big
Deals have gained support and remain firmly in place in the form of
transformative agreements. However, in addition to research funders and
libraries that increasingly make publishing contracts public, it is on
the publisher to provide metadata as presented here to improve the
transparency in scholarly publishing by linking invoicing data to
bibliometrics. From this study we can affirm that hybrid OA is complex
as the financial flows come from various stakeholders and invoicing
forms, it is not just libraries or consortia, but authors and funders
are also directly involved. As of right now, providing machine-readable
article-level invoicing data is not common amongst publishers, which
limits the potential for broader studies.

\hypertarget{data-availability}{%
\section{Data Availability}\label{data-availability}}

All data and code used for this study is openly available on GitHub,
including version history:

\url{https://github.com/njahn82/elsevier_hybrid_invoicing}

Formal analyis is written in R Markdown and thus replicable.

\hypertarget{references}{%
\section*{References}\label{references}}
\addcontentsline{toc}{section}{References}

@@

\leavevmode\hypertarget{ref-Bj_rk_2010}{}%
Björk, B.-C., Welling, P., Laakso, M., Majlender, P., Hedlund, T., \&
Guðnason, G. (2010). Open access to the scientific journal literature:
Situation 2009. \emph{PLoS ONE}, \emph{5}(6), e11273.
\url{https://doi.org/10.1371/journal.pone.0011273}

@@

\leavevmode\hypertarget{ref-Castro_2019}{}%
Castro, P. de. (2019). \emph{Running a no-hybrid open access funding
policy: Some results}. Open Access Strathclyde.
\url{https://strathoa.tumblr.com/post/185703709380/running-a-no-hybrid-open-access-funding-policy}

@@

\leavevmode\hypertarget{ref-Chawla_2020}{}%
Chawla, D. (2020). This tool is saving universities millions of dollars
in journal subscriptions. \emph{Science}.
\url{https://doi.org/10.1126/science.abd7483}

\leavevmode\hypertarget{ref-Creaser_2010}{}%
Creaser, C., Fry, J., Greenwood, H., Oppenheim, C., Probets, S., Spezi,
V., \& White, S. (2010). Authors' awareness and attitudes toward open
access repositories. \emph{New Review of Academic Librarianship},
\emph{16}(sup1), 145--161.
\url{https://doi.org/10.1080/13614533.2010.518851}

@@

\leavevmode\hypertarget{ref-Fraser_2020}{}%
Fraser, N., Brierley, L., Dey, G., Polka, J. K., Pálfy, M., Nanni, F.,
\& Coates, J. A. (2020). \emph{Preprinting the COVID-19 pandemic}.
bioRxiv. \url{https://doi.org/10.1101/2020.05.22.111294}

@@

\leavevmode\hypertarget{ref-Huang_2020}{}%
Huang, C.-K. (Karl), Neylon, C., Hosking, R., Montgomery, L., Wilson, K.
S., Ozaygen, A., \& Brookes-Kenworthy, C. (2020). Evaluating the impact
of open access policies on research institutions. \emph{eLife},
\emph{9}. \url{https://doi.org/10.7554/elife.57067}

@@

\leavevmode\hypertarget{ref-Jubb_2017}{}%
Jubb, M., Plume, A., Oeben, S., Brammer, L., Johnson, R., Bütün, C., \&
Pinfield, S. (2017). \emph{Monitoring the transition to open access:
December 2017}.
\url{https://www.universitiesuk.ac.uk/policy-and-analysis/reports/Documents/2017/monitoring-transition-open-access-2017.pdf}

@@

\leavevmode\hypertarget{ref-Kramer_2018}{}%
Kramer, B., \& Bosman, J. (2018). \emph{Towards a plan S gap analysis:
open access potential across disciplines using Web of Science and DOAJ}
{[}Data set{]}. Zenodo. \url{https://doi.org/10.5281/zenodo.1979937}

@@

\leavevmode\hypertarget{ref-Van_Noorden_2013}{}%
Noorden, R. V. (2013). Researchers opt to limit uses of open-access
publications. \emph{Nature}.
\url{https://doi.org/10.1038/nature.2013.12384}

@@

\leavevmode\hypertarget{ref-Pieper_2018}{}%
Pieper, D., \& Broschinski, C. (2018). OpenAPC: A contribution to a
transparent and reproducible monitoring of fee-based open access
publishing across institutions and nations. \emph{Insights the UKSG
Journal}, \emph{31}. \url{https://doi.org/10.1629/uksg.439}

@@

\leavevmode\hypertarget{ref-Robinson_Garcia_2020}{}%
Robinson-Garcia, N., Costas, R., \& Leeuwen, T. N. van. (2020). Open
access uptake by universities worldwide. \emph{PeerJ}, \emph{8}, e9410.
\url{https://doi.org/10.7717/peerj.9410}

\leavevmode\hypertarget{ref-Rowley_2017}{}%
Rowley, J., Johnson, F., Sbaffi, L., Frass, W., \& Devine, E. (2017).
Academics\\
textquotesingle behaviors and attitudes towards open access publishing
in scholarly journals. \emph{Journal of the Association for Information
Science and Technology}, \emph{68}(5), 1201--1211.
\url{https://doi.org/10.1002/asi.23710}

\leavevmode\hypertarget{ref-Schimmer_2015}{}%
Schimmer, R., Geschuhn, K., \& Vogler, A. (2015). \emph{Disrupting the
subscription journals'business model for the necessary large-scale
transformation to open access}. Max Planck Digital Library.
\url{https://doi.org/10.17617/1.3}

\leavevmode\hypertarget{ref-Severin_2020}{}%
Severin, A., Egger, M., Eve, M. P., \& Hürlimann, D. (2020).
Discipline-specific open access publishing practices and barriers to
change: An evidence-based review. \emph{F1000Research}, \emph{7}.
\url{https://doi.org/10.12688/f1000research.17328.2}

@@

\leavevmode\hypertarget{ref-Tickel_2018}{}%
Tickel, A. (2018). \emph{Open access to research publications -- 2018}.
\url{https://assets.publishing.service.gov.uk/government/uploads/system/uploads/attachment_data/file/774956/Open-access-to-research-publications-2018.pdf}

\leavevmode\hypertarget{ref-birmingham}{}%
University of Birmingham. (n.d.). \emph{UKRI open access block grant}.
\url{https://intranet.birmingham.ac.uk/as/libraryservices/library/research/open-access/funding/ukri-open-access-block-grant.aspx}

\leavevmode\hypertarget{ref-oxford_2019}{}%
University of Oxford. (2019). \emph{Change to oxford's policy for rcuk
oa block grant 2nd december 2019}.
\url{http://openaccess.ox.ac.uk/2019/12/02/change-to-oxfords-policy-for-rcuk-oa-block-grant-2nd-december-2019/}
